Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-08 aistats.tex
Complete quadratic comparison

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

\newcommand{\MAPm}{\hat \mu_n}
\newcommand{\MAPt}{\hat \natp_n}

@@

\subsection{Quadratic Regime}
As another reference point, let us consider the case $\logpart = \half \norm{\natp}_2^2$, which corresponds to finding the mean of a Gaussian with variance $1$ ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
If we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
    &\leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.

@@

@@

\subsection{Gaussian Mean}
\paragraph{$\ell^2$-norm Analysis}
In the Euclidean realm, the average is bounded by variance becomes a bound on the variance of the average
\begin{align}
	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
\end{align}
eg the variance of the mean is $n$ times smaller than the variance of the samples.
Adding a reference mean $\mu_0$ to get the MAP yields
\begin{align}
	&\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
	\\
	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
\end{align}
so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !

>>>>

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\cond}{\,\vert\,}
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\nat}{\theta}
\newcommand{\natp}{\theta}
\newcommand{\m}{m}
\newcommand{\meanp}{\m}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

\newcommand{\MAPm}{\hat \m_n}
\newcommand{\MAPt}{\hat \natp_n}

@@

\subsection{Quadratic Case}
As another reference point, let us consider the case $\logpart(\natp) = \half \norm{\natp}_2^2$.
For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples we can break down the MLE into individual point's contributions
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
	=\frac{\Var(T)}{2n} 
\end{align}
Adding a reference mean $\mu_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\mu^* -  \mu_0}^2}{(n+n_0)^2}
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
\end{align}
so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
    &\leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.

@@

\subsection{Locally Quadratic Case}
All Bregman divergences are locally quadratic. 
Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
-- e.g. exponential distribution, 
Pareto distribution with known minimum value, 
Weibull distribution with known shape k, 
Laplace distribution
with known mean
-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
\begin{align}
	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
\end{align}
holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\natp^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.

\TODO{Reference \citet{ostrovskii2021finite}.}

All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.

@@
